Seeq: a library for inexact DNA sequence matching

### 2015-04-01 doc/document.tex
Updated manuscript.

\section{Motivation:}
Searching specific sequences in DNA is one of the most common tasks
in the analysis of biological data. Experimental samples often contain
known sequences at the positions where adaptor removal, sequence
trimming or barcode extraction have to be performed. Experimental data
contain errors that must be tolerated in order to identify these
patterns. For reads produced by second generation sequencing,
where the error rates lie below 2\%, short sequences with small error
tolerance suffice to identify the patterns. However, third generation
sequencing technologies promise longer reads at the cost of 10-fold
higher error rates. In this scenario, both the pattern length and the
relative error tolerance have to be dramatically increased.
\section{Results:}
Here we present seeq, an algorithm for pattern matching with errors.
Seeq is optimized for DNA sequences and shows very little increase in
running time as error tolerance increases. We benchmark seeq against
the most widely used algorithms for inexact pattern matching. The
results show that seeq scales better and becomes significantly faster
as the error rate increases.
\section{Availability:}
Seeq is available in three forms: a Linux software, a C library and a 
Python module. The source code and the C/Python libraries are available
for download at
\href{http://github.com/ezorita/seeq}{http://github.com/ezorita/seeq}.
\section{Contact:} \href{guillaume.filion@gmail.com}{guillaume.filion@gmail.com}
\end{abstract}

@@

Inexact string matching dates back to the origin of computer sciences
\citep{Lev65}. The first algorithms addressing this problem were based
on dynamic programming \citep{NW70}. They are still widely used to
align biological sequences due to their simplicity and flexibility,
More sophisticated algorithms derived from the Boyer-Moore search
method \citep{BM77} were later designed to search patterns on long
strings allowing errors. Such algorithms are faster than the dynamic
programming approach because they do not compute all the relative
distances. Instead, the pattern is pre-processed to design a search
strategy which skips parts of the text that do not contain any
match. The jumping strategy is very efficient for perfect matches or
small edit distances because non-matching conditions can be inferred
with only a few comparisons. In those conditions, the number of
operations per input character is small because large areas of the
text are never processed. However, when the error rate is high, the
search strategy requires more comparisons to identify a non-matching
condition. The drop in performance is even more accute when the alphabet
is as small as in the case of DNA sequences.

@@

Here we present \emph{seeq}, an algorithm using a dynamic version of
deterministic finite automata (DFA). Instead of using standard
construction procedures, the dynamic DFA of seeq is built as the input
text is processed. We benchmark seeq against other widely used inexact
string matching software. The results show that seeq scales better and
becomes significantly faster as the error rate increases.

\begin{methods}
\section{Methods}

@@

\begin{figure}[!tpb]
%\centerline{\includegraphics[scale=0.5]{fig_methods.pdf}}
\caption{}\label{matrix}
\end{figure}

The DFA construction process is illustrated in Figure. X. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{enumerate}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then set $S = S_q$.
\item Otherwise, compute the column of the edit matrix corresponding
to $f(S, T_j)$. Case 2a: if this state already exists ($S_q$), connect
$S$ to it with a path labeled $T_j$ and set $S = S_q$. Case 2b: if
the state does not exist, create it, and do as in in the previous
case 2a.
\end{enumerate}

@@

In practice, most pattern searches in DNA reads are performed from
scripts and pipelines. To ease the use of seeq in existing software,
we have ensured that it is comliant with Linux pipes. Also, we have
created a Python module that can be imported and integrated seamlessly,
as it can be called on Python string objects that can be further
manipulated. We also provide a API in C for the cases that optimal
performance is required.

>>>>

\section{Motivation:}
Searching specific sequences in DNA is one of the most common tasks
in the analysis of biological data. Experimental samples often contain
known sequences at the positions where adaptor removal, sequence
trimming or barcode extraction have to be performed. Experimental data
contain errors that must be tolerated in order to identify these
patterns. For reads produced by second generation sequencing,
where the error rates lie below 2\%, short sequences with small error
tolerance suffice to identify the patterns. However, third generation
sequencing technologies promise longer reads at the cost of 10-fold
higher error rates. In this scenario, both the pattern length and the
relative error tolerance have to be dramatically increased.
\section{Results:}
Here we present seeq, an algorithm for pattern matching with errors.
Seeq is optimized for DNA sequences and shows very little variance in
running time with different configurations. We benchmark seeq against
the most widely used algorithms for inexact pattern matching. The
results show that seeq scales better and becomes significantly faster
as the error rate increases.
\section{Availability:}
Seeq is available in three forms: a Linux software, a C library and a 
Python module. The source code and the C/Python libraries are available
for download at
\href{http://github.com/ezorita/seeq}{http://github.com/ezorita/seeq}.
\section{Contact:} \href{guillaume.filion@gmail.com}{guillaume.filion@gmail.com}
\end{abstract}

@@

Inexact string matching dates back to the origin of computer sciences
\citep{Lev65}. The first algorithms addressing this problem were based
on dynamic programming \citep{NW70}. They are still widely used to
align biological sequences due to their simplicity and flexibility.
More sophisticated algorithms derived from the Boyer-Moore search
method \citep{BM77} were later designed to search patterns on long
strings allowing errors. Such algorithms are faster than the dynamic
programming approach because they do not compute all the relative
distances. Instead, the pattern is pre-processed to design a search
strategy which skips parts of the text that do not contain any
match. The jumping strategy is very efficient for perfect matches or
small edit distances because non-matching conditions can be inferred
with only a few comparisons. In such conditions, the number of
operations per input character is small because large areas of the
text are never processed. However, when the error rate is high, the
search strategy requires more comparisons to identify a non-matching
condition. The drop in performance is even more accute when the alphabet
is as small as in the case of DNA sequences.

@@

Here we present \emph{seeq}, an algorithm using a dynamic version of
deterministic finite automata (DFA). Instead of using standard
construction procedures, the dynamic DFA of seeq is built as the input
text is processed. We benchmark seeq against other inexact string
matching software. The results show that seeq scales better and
becomes significantly faster as the error rate increases. 
\begin{methods}
\section{Methods}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.6]{Figure_1.pdf}}
\caption{Dynamic DFA construction process for T='TGTGCA' and P='GCAC'.
a) Dynamic programming matrix of the alignment. The differential code
of each column is shown inside the semicircles. b) State of the DFA at
the end of the alignment.}\label{matrix}
\end{figure}

The DFA construction process is illustrated in Figure. \ref{matrix}. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{enumerate}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then set $S = S_q$.
\item Otherwise, revert $\mathbf{c}$ from $\mathbf{s}$ and compute
the $j$-th column of the edit matrix as $\mathbf{c}_j = f(\mathbf{c}, T_j)$ and
its diferential code $\mathbf{s}_j$. Then: a) If $\mathbf{s}_j$
represents an existing state ($S_q$), connect $S$ to it with a path
labeled $T_j$ and set $S = S_q$. b) If none of the states is
represented by $\mathbf{s}_j$, create $S_j$, connect $S$ to it with a
path labeled $T_j$ and set $S = S_j$.
\end{enumerate}

@@

In practice, most pattern searches in DNA reads are performed from
scripts and pipelines. To ease the use of seeq in existing software,
we have ensured that it is comliant with Linux pipes. Also, we have
created a Python module that can be imported and integrated seamlessly,
as it can be called on Python string objects that can be further
manipulated. We also provide an API in C for the cases that optimal
performance is required.
